### 2009-11-20 docs/marcello-dissertation/thesis/chapter2-state-of-the-art.tex
Revision of the chapters 2-6 regarding Data Provenance in sensor networks:
* Chapter 2
  - Adding section regarding provenance

* Chapter 3
  - Splitting section 1 into 2 different ones; TODO: Draw another diagram. 
  - Adding provenance information to the taxonomy of data model
  
* Chapter 4
  - Adding references regarding to the missing location in the NetBEAMS data;

* Chapter 5
  - References to the provenance sections to the keys definition;
  
* Chapter 6
  - References to the provenance sections to the keys definition;

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence

\begin{itemize}
  \item \textbf{Tabular Data Model}: comma-separeted values are very popular
  due to its simplicity. Although data is not index, it is an easy approach for data
 exchange among researchers.
  \item \textbf{Relational Data Model}: it is used in the majority of the
  projects. Most of them due to the easy access to database systems, with
  refactored versions that support the implementation of sensor networks and
  its requirements \cite{sn-db-tinydb};
  \item \textbf{Structured Data Model}: simiar to the relational model,
  structured data models are becoming popular due to the Internet. The use of
  XML data \cite{xml} is not yet popular, but there are a few projects using it
  \cite{sn-xml-usage01}\cite{sn-xml-usage01}, implementing its query by using a
  middleware \cite{sn-xml-middleware} \cite{sn-xml-query-engines}.
  \item \textbf{Key-Value Pair and Document-Oriented Data Model}: as
  an emerging data model, the Key-Value Pair (KVP) \cite{db-kvp} data model is
  becoming popular with the up-coming trend of cloud computing. Although there are no
  records of the use of this model data model in sensor networks, this work
  will investigate it.
\end{itemize}

Most importantly, the use of Data Provenance approaches to describe the
collected data is shown to be important \cite{sn-provenance}, given the nature
of the collected data. Therefore, this section describes these approaches for
modeling data collected from sensor devices, taking into account the need of
metadata for the data.

@@

\subsection{Key-Value Pair and Document-Oriented Data Models}

The advance of distributed systems and the Internet have enabled the
development of powerful database systems that can be organized in the context
of database systems and how to represent models. An important up-coming
approach is the use of database systems that does not use structured language
for quering its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}. Such model has the following properties:

@@

\subsection{Data Provenance for Sensor Networks}

Most of sensor networks implementations use the collected data from the sensor
devices after they observe the environment. As shown by
\cite{sn-data-model-survey}, different data models contains its pros and
cons to represent data. However, one important aspect in sensor networks is the
description of it collected data, as \cite{sn-provenance} discussed the
properties and nature of sensor networks generated data. First, the device's
sampling are usually in raw format or without obvious names. In order to
provide search over the collected sensor data, the data needs first to be
described and indexed, and this process can be related to Data Provenance
\cite{db-provenance}.

@@

\begin{itemize}
  \item each low-level observation as it describes the actual data;
  \item the location where the collected data was generated;
  \item the time when the data was collected.
\end{itemize}

While Data Provenance provides necessary mechanisms to describe sensor networks
data in general, others see the importance of the time space of sensor
networks. \cite{sn-time-series} describes strategies to used to cluster
distributed sensor data as time series of data.

>>>>

\begin{itemize}
  \item \textbf{Tabular Data Model}: comma-separeted values are very popular
  due to its simplicity. Although data is not index, it is an easy approach for data
 exchange among researchers \cite{sn-provenance}.
  \item \textbf{Relational Data Model}: it is used in the majority of the
  projects. Most of them due to the easy access to database systems, with
  refactored versions that support the implementation of sensor networks and
  its requirements \cite{sn-db-tinydb};
  \item \textbf{Structured Data Model}: simiar to the relational model,
  structured data models are becoming popular due to the Internet. The use of
  XML data \cite{xml} is not yet popular, but there are a few projects using it
  \cite{sn-xml-usage01}\cite{sn-xml-usage01}, implementing its query by using a
  middleware \cite{sn-xml-middleware} \cite{sn-xml-query-engines}.
\end{itemize}

Although these approaches have been used in Sensor Networks, this work will
also take a look at others still not used, as follows:

\begin{itemize}
  \item \textbf{Key-Value Pair Model}: as an emerging data model, the Key-Value
  Pair (KVP) \cite{db-kvp} data model is becoming popular with the up-coming
  trend of cloud computing. Although there are no records of the use of this
  model data model in sensor networks, this work will investigate it.
\end{itemize}

@@

\subsection{Key-Value Pair Data Models}

The advance of distributed systems and the Internet have enabled the
development of powerful database systems that can be organized in the context
of database systems and how to represent models. An important up-coming
approach is the use of database systems that does not use structured language
for quering its data. One such example is called Key-Value Pair (KVP) data model
\cite{db-kvp}, which is also known as Document-Oriented, Distributed Hash Table.
Such model has the following properties:

@@

\section{How to describe collected sensor data using Data Provenance}
\label{sec:sn-provenance}

Most of sensor networks implementations use the collected data from the sensor
devices after they observe the environment. As shown by
\cite{sn-data-model-survey}, different data models contains its pros and
cons to represent data. However, one important aspect in sensor networks is the
description of it collected data, as \cite{sn-provenance} discussed the
properties and nature of sensor networks generated data. First, the device's
sampling are usually in raw format or without obvious names. In order to
provide search over the collected sensor data, the data needs first to be
described and indexed, and this process can be related to Data Provenance
\cite{db-provenance}. Needless to say, Data Provenance is the 
Metadata\footnote{Data about data} about the collected sensor data.

@@

\begin{itemize}
  \item \textbf{what was collected?}: it can be simply one or more low-level
  observations, which is a set of values with or without a description;
  \item \textbf{When was the data collected?}: is the the time when the data was
  collected, as most sensor data must be used depending on its existance;
  \item \textbf{Where was the data collected?}: it is primarily the location
  where the collected data was observed. However, It is important to note that
  the use of this metadata depends on the applicability of the data.
\end{itemize}

While Data Provenance provides necessary mechanisms to describe sensor networks
data in general, others see the importance of the time space of sensor
networks. \cite{sn-time-series} describes strategies to used to cluster
distributed sensor data as time series of data, as well as to decrease the size
of data streams being transmitted in real-time sensor networks when these
collected data does not need to be transferred \cite{sn-data-reduction}.

Finally, an important application of collected sensor data is the so-called
blackbox of an aircraft, which carries the collection of raw data collected
from a group of sensors placed in different places of the aircraft, as
described by \cite{sn-exemple-blackbox}. The author describes the data is
used in a process of data recovery and analysis in order to identify the
possible reason of an aircraft accident, by using the time, and location of
the aircraft, along with the different variables from the sensors.

\subsection{What was collected?}

Different sensor devices will provide different attribute sets regarding a
specific observation. In general, these data sets does not have obvious names,
which leads to difficulties when indicies are necessary in order to make the
data searcheable. Specially for the purpose of data archival, as described
earlier in this chapter, the collected data needs to receive and identity,
which identifies the data by itself in a regular database
\cite{relational-model}. Equally, \cite{sn-provenance} described that
traditional sensor networks uses the collected data for analysis long after
the attribute set has been collected, usually before it is categorized and
indexed. Annotations and tags are also names given by \cite{sn-provenance} as
means to describe collected sensor data.

In a nutshell, data is described by the name of the observation, when this is
not provided. For example,

\begin{itemize}
  \item \textbf{temperature} = 73.45: this is the value of the collected
  observation, which in general, the attribute name ``temperature'' is not
  provided in the data stream;
  \item \textbf{temperature-scale} = ``Fahrenheit'': the metadata regarding the
  temperature attribute, which describes the scale of the temperature.
  \item \textbf{device-version} = 5: the metadata that describes the version of
  the sensor device, since sensors can be replaced, and as a consequence, data
  accuracy may be as well be changed.
\end{itemize}

\subsection{When was the data collected?}

In order to use historical sensor data, the use of a point in time is the most
important characteristic of data, as it is considered an important metadata
for different types of sensor networks such as an aircraft the blackbox
\cite{sn-exemple-blackbox}. Other projects which classifies data in a
time-series \cite{sn-time-series-example}, while others provide approaches to
decrease the creation of data based on clustering data before they are sent to
the sink \cite{sn-time-series}.

In different areas, temporal databases or time-series are the ones that takes
the real-world time into consideration. In general, data on a temporal
database can be classified by its dimention, as surveyed by \cite{db-temporal}:

\begin{itemize}
  \item \textbf{Valid time}: it is the point in time during which a fact
  happened with respect to the real world. This value is only when data is captured, an
  cannot be changed because of its nature, as well as placing a restriction over
  Update operations;
  \item \textbf{Transaction time}: is the point in time at which a fact is
  collected, transferred or processed in the database. This value is used
  during Insertion operations, as well as in Update ones.
\end{itemize}

When both temporal dimensions are applied, the data is consided to be 
Bitemporal when \cite{db-temporal}. Usually, the data type of the valid and
transaction time columns are simple timestamps, with any supported accuracy.
For example, the following set of attribute and values shows a temperature of
70 degrees that was observed on November 15, 2009, but collected only 3 days
later.

\begin{itemize}
  \item \textbf{temperature} = 70;
  \item \textbf{fact-time} = Nov 15, 2009 10:45 PM PST
  \item \textbf{transaction-time} = Nov 18, 2009 11:11 PM PST
\end{itemize}

When it comes to Data Provenance in sensor networks, \cite{sn-provenance}
shows that the collected data for historical purpose is usually saved in
self-described flat-files by file-names, containing complicated naming
conventions that includes the timestamp information about the data. For this
reason, time is an important metadata needed in any data model used, since it 
helps identifying data uniquely for a given data set.

\subsection{Where was the data collected from?}

Depending on the application of the collected sensor data, the location in
which the data was collected may be needed. For instance, the geographic
location of the collected data is important to identify the specific
GPS\footnote{Global Positioning System} coordinates of the sensor device
\cite{sn-ex02}. Similarly, the position of any event captured in a black-box of
an aircraft carries is also used as a metadata for the event. It will either
depend on the sensor device provide such sampling attribute, or entered
manually at the database.

A correlation between time and location is presented by \cite{sn-time-series}
as an important feature to reduce the traffic generated in a sensor network by
clustering similar data when the sensor devices are placed in concentrated
environment.

### 2009-11-20 docs/marcello-dissertation/thesis/chapter3-taxonomic-classification-of-persistence-storage-for-sensor-networks.tex
Revision of the chapters 2-6 regarding Data Provenance in sensor networks:
* Chapter 2
  - Adding section regarding provenance

* Chapter 3
  - Splitting section 1 into 2 different ones; TODO: Draw another diagram. 
  - Adding provenance information to the taxonomy of data model
  
* Chapter 4
  - Adding references regarding to the missing location in the NetBEAMS data;

* Chapter 5
  - References to the provenance sections to the keys definition;
  
* Chapter 6
  - References to the provenance sections to the keys definition;

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence

\section{Nature and Location of Sensor Data Taxonomy}

@@

\begin{itemize}
  \item \textbf{Real-time Data Stream}: data is used as real-time data stream
  during data query \cite{sn-intro01}.
    \begin{itemize}[label=\textbullet]
        \item \textbf{Local Storage}: The data can be temporarily persisted
        in-memory, or a secondary memory device in a local storage device.
    \end{itemize}
  \item \textbf{Archival Data}: data is used for historical purposes such as
  data analysis or data fusion \cite{sn-intro01, sn-intro02}.  
    \begin{itemize}[label=\textbullet]
      \item \textbf{External Storage}: The data is usually located in an
       external storage device with a data management system such as a database
       system.
      \item \textbf{Data-Centric Storage}: the collected data can arranged in a
      categorized fashion as described by \cite{sn-storage03}. This taxon of
      system allows the data to be partitioned by a specific property. Nowadays,
      the strategy is often related to database partition \cite{db-partition}
      and database sharding approaches \cite{db-shard01} \cite{db-shard02}.
    \end{itemize}
\end{itemize}

@@

>>>>

\section{Nature of Sensor Data Taxonomy}

@@

\begin{itemize}
  \item \textbf{Real-time Data Stream}: data is used as real-time data stream
  during data query \cite{sn-intro01};
  \item \textbf{Archival Data}: data is used for historical purposes such as
  data analysis or data fusion \cite{sn-intro01, sn-intro02}.
\end{itemize}

\section{Location of Sensor Data Taxonomy}

\begin{itemize}
  \item \textbf{Local Storage}: The data can be temporarily persisted
   in-memory, or a secondary memory device in a local storage device. It is
  directly related to the real-time data stream taxon, since it occurs at
  the place of the sensor device;
  \item \textbf{External Storage}: The data is usually located in an
  external storage device with a data management system such as a database
  system.
  \item \textbf{Data-Centric Storage}: the collected data can arranged in a
  categorized fashion as described by \cite{sn-storage03}. This taxon of
  system allows the data to be partitioned by a specific property. Nowadays,
  the strategy is often related to database partition \cite{db-partition}
  and database sharding approaches \cite{db-shard01} \cite{db-shard02}.
\end{itemize}

@@

No matter the type of data model used, the support for Data Provenance as shown
in section \ref{sec:sn-provenance} is as important as the data being
collected. For this reason, the data model taxonomy must have support for the
data types that support the use of provenance-related attributes in the
collected sensor data.

### 2009-11-20 docs/marcello-dissertation/thesis/chapter4-netbeams-in-depth-requirements-analysis-and-specification.tex
Revision of the chapters 2-6 regarding Data Provenance in sensor networks:
* Chapter 2
  - Adding section regarding provenance

* Chapter 3
  - Splitting section 1 into 2 different ones; TODO: Draw another diagram. 
  - Adding provenance information to the taxonomy of data model
  
* Chapter 4
  - Adding references regarding to the missing location in the NetBEAMS data;

* Chapter 5
  - References to the provenance sections to the keys definition;
  
* Chapter 6
  - References to the provenance sections to the keys definition;

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence

\chapter{Netbeams in Depth: Requirements Analysis and Specification}

@@

\begin{itemize}
  \item 5 YSI sondes in operation at the RTC site Pier in Tiburon Island, San
  Francisco Bay;
  \item Each YSI sonde produces 52 Bytes for each its observations;
  \item The sampling frequency rate could be configured in ranges of 1, 6 or
  15 minutes, depending on the staff's speficiation. 
\end{itemize}

@@

\begin{itemize}
  \item Main users are biologists, no expertise in Database Systems;
  \item Data Representation in human-readable format;
  \item Scalable system with low degree of maintanance;
  \item Data searchable in near-real-time
\end{itemize}

@@

\begin{table}
    \caption{DSP Matcher Rule Algorithm to filter Data Consumer components to
    receive a given Message}
    \begin{center}
        \begin{tabular}{lr}
          \textit{Data Consumers ( DSP Message ) := Verify the DSP
          Message's Header + Verify the matcher rules (which contains the list
          of consumers)}
        \end{tabular}
    \end{center}
    \label{tab:ysi-data-stream}
\end{table}

>>>>

\chapter{Netbeams in Depth: Requirements Analysis and Specification}
\label{chap:netbeams-overview}

@@

\begin{itemize}
  \item 5 YSI sondes in operation at the RTC site Pier in Tiburon Island, San
  Francisco Bay;
  \item Each YSI sonde produces 52 Bytes for each its observations;
  \item The sampling frequency rate could be configured in ranges of 1, 6 or
  15 minutes, depending on the staff's speficiation;
  \item Data for final users are published with Provenance Data regarding time,
  but not regarding location.
\end{itemize}

@@

\begin{itemize}
  \item Main users are biologists, no expertise in Database Systems;
  \item Data Representation in human-readable format;
  \item Scalable system with low degree of maintanance;
  \item Data searchable in near-real-time.
\end{itemize}

@@

\begin{table}
    \caption{DSP Matcher Rule Algorithm to filter Data Consumer components to
    receive a given Message}
    \begin{center}
        \begin{tabular}{lr}
          \textit{Data Consumers ( DSP Message )} := \\ 
          Verify the DSP Message's Header + \\
          Verify the matcher rules, which contains the list of consumers for
          the given message
        \end{tabular}
    \end{center}
    \label{tab:ysi-data-stream}
\end{table}

### 2009-11-20 docs/marcello-dissertation/thesis/chapter5-taxonomic-persistence-storage-selection-for-netbeams.tex
Revision of the chapters 2-6 regarding Data Provenance in sensor networks:
* Chapter 2
  - Adding section regarding provenance

* Chapter 3
  - Splitting section 1 into 2 different ones; TODO: Draw another diagram. 
  - Adding provenance information to the taxonomy of data model
  
* Chapter 4
  - Adding references regarding to the missing location in the NetBEAMS data;

* Chapter 5
  - References to the provenance sections to the keys definition;
  
* Chapter 6
  - References to the provenance sections to the keys definition;

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence

For instance, all data needed for the YSI sonde data, as well as all necessary
provenance data that describes the data, are represented by means of key-value
pairs \ref{fig:persistence-example-kvp}.

@@

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enable them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

>>>>

For instance, all data needed for the YSI sonde data, as well as all necessary
provenance data that describes the data, depicted in section
\ref{sec:sn-provenance}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

@@

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships;
      \item Data Provenance can be provided through data types such as
      ``Timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enable them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

### 2009-11-20 docs/marcello-dissertation/thesis/chapter6-data-persistence-component-design-and-architecture.tex
Revision of the chapters 2-6 regarding Data Provenance in sensor networks:
* Chapter 2
  - Adding section regarding provenance

* Chapter 3
  - Splitting section 1 into 2 different ones; TODO: Draw another diagram. 
  - Adding provenance information to the taxonomy of data model
  
* Chapter 4
  - Adding references regarding to the missing location in the NetBEAMS data;

* Chapter 5
  - References to the provenance sections to the keys definition;
  
* Chapter 6
  - References to the provenance sections to the keys definition;

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence

After analyzing the DSP Platform and the structure of the DSP Messages in
chapter \ref{netbeams-architecture}, the following set of keys are expected to
be part of the design of the data model:

>>>>

After analyzing the DSP Platform and the structure of the DSP Messages in
chapter \ref{netbeams-architecture}, the following set of keys are expected to
be part of the design of the data model. Any key represented by time or
location are related to the provenance support discussed in section
\ref{sec:sn-provenance}:
